Computing Entropies With Nested Sampling

### 2017-08-14 paper/ms.tex
Final changes

From Equations~\ref{eqn:minuslogX} and~\ref{eqn:exp}, we see that}
the sequence of $-\ln X$ values produced by a Nested Sampling run can be
considered as a Poisson process with rate equal to $N$, the number of
particles. This is why separate NS runs can be simply merged
\citep{skilling2006nested}.
\citet{Walter2017point} showed how this view of the NS sequence of points
can be used to construct a
version of NS that produces unbiased (in the frequentist sense)
estimates of the evidence.
However, it can also be used to construct an unbiased estimator of
a log-probability, which is more relevant to information theoretic
quantities discussed in the present paper.

@@

If we actually want the differential entropy
\begin{align}
H(\x) &= -\int p(\x) \log p(\x) \, d\x
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $r$ is small, so that
\begin{align}
P(d(\x; \xref) < r \,|\, \xref)
    &\approx
    p(\xref) \int_{d(\x; \xref) < r} \, d\x\\
    &= p(\xref) \, V
\end{align}
where $V$ is the volume of the region where $d(\x; \xref) < r$.
Then $H(\x) = \left< \depth \right>_{p(\x)} + \log V(r)$.
{\revision It is important to remember that the definition of
volume here is in terms of the coordinates $\x$, and that a change
of coordinates (or a change of the definition of the distance function)
would result in a different value of the
differential entropy.}

@@

To use the algorithm, a distance function $d(;)$ must be defined, to quantify
how far the NS particles are from the reference particle. To calculate
the entropy of the data, {\revision I applied the algorithm in the joint
space of possible parameters and datasets}, i.e., distribution of reference
particles followed Equation~\ref{eqn:joint_prior}.
For the distance function, which helps the NS particles approach the
reference particle, I used a
simple Euclidean metric in the space of datasets:
\begin{align}
d\left(\x; \xref\right) &= \sqrt{
        \sum_{i=1}^{100}\left(x_i - x_{{\rm ref}, i}\right)^2}.
\end{align}
Since the distance function refers only to the data $\boldsymbol{x}$
and not the parameter $\mu$, the sampling is effectively done in the
space of possible datasets only --- the parameter $\mu$ can be thought of
as merely a
latent variable that allows $p(\boldsymbol{x})$ to be explored.
}

@@

However, more significantly, the above method
would not work if we want $I(\x; \d)$ for
a single parameter or a subset of them, rather than for all of the parameters.
{\revision
The prior-to-posterior KL divergence measured by standard NS is the
amount learned about {\em all} parameters, including nuisance parameters}.
Therefore, I adopted the third formula as the best way to compute $I(\x; \d)$.
In the sinusoidal example, $H(\x)$ is available analytically, and I use the algorithm to
obtain $H(\x | \d)$ as follows:
\begin{align}
H(\x | \d) &= \int p(\d_{\rm ref}) H(\x | \d=\d_{\rm ref}) \, d\d_{\rm ref}.
\end{align}
This is an expected value over datasets, of the entropy of the posterior given
each dataset. I ran the algorithm by generating reference particles from
$p(\x, \d)$, i.e., generating parameter values from the prior, along with
corresponding simulated datasets. Given each simulated dataset, I used NS
to compute an unbiased estimate of $H(\x | \d=\d_{\rm ref})$, by evolving
a set of particles (initially from the posterior) towards the reference particle.
The role of the ``prior'' in the NS is actually taken by the posterior, and the
``likelihood'' was minus a distance function defined in terms of the
parameter(s) of interest.

@@

Letting $\tau = \log_{10} T$, and treating
$A$ and $\phi$ are nuisance parameters, I calculated the conditional entropy
$H(\tau | \boldsymbol{Y})$ using the algorithm with a tolerance of $r = 10^{-5}$.
The result was
\begin{align}
\left<\depth\right> &= 5.379 \pm 0.038
\end{align}
Converting to a differential entropy by adding {\revision $\ln (2r) = -10.820$}, the conditional
entropy is
\begin{align}
H(\tau | \boldsymbol{Y}) &= -5.441 \pm 0.038 \textnormal{ nats}.
\end{align}
Since the Uniform(0,1) prior for $\tau$ has a differential entropy of zero, the
mutual information is $I(\tau; \boldsymbol{Y}) = H(\tau) - H(\tau | \boldsymbol{Y}) = 5.441 \pm 0.038$
nats.

@@

>>>>

From Equations~\ref{eqn:minuslogX} and~\ref{eqn:exp}, we see that}
the sequence of $-\ln X$ values produced by a Nested Sampling run can be
considered as a Poisson process with rate equal to $N$, the number of
particles. This is why separate NS runs can be simply merged
\citep{skilling2006nested, HENDERSON201784}.
\citet{Walter2017point} showed how this view of the NS sequence of points
can be used to construct a
version of NS that produces unbiased (in the frequentist sense)
estimates of the evidence.
However, it can also be used to construct an unbiased estimator of
a log-probability, which is more relevant to information theoretic
quantities discussed in the present paper.

@@

If we actually want the differential entropy
\begin{align}
H(\x) &= -\int p(\x) \log p(\x) \, d\x
\end{align}
we can use the fact that density equals mass divided
by volume. Assume $r$ is small, so that
\begin{align}
P(d(\x; \xref) < r \,|\, \xref)
    &\approx
    p(\xref) \int_{d(\x; \xref) < r} \, d\x\\
    &= p(\xref) \, V
\end{align}
where $V$ is the volume of the region where $d(\x; \xref) < r$.
Then $H(\x) = \left< \depth \right>_{p(\x)} + \log V(r)$.
{\revision It is important to remember that the definition of
volume here is in terms of the coordinates $\x$, and that a change
of coordinates (or a change of the definition of the distance function)
would result in a different value of the
differential entropy. This is equivalent to counting possibilities
or volumes with a different choice of base measure $q(\x)$.}

@@

To use the algorithm, a distance function $d(;)$ must be defined, to quantify
how far the NS particles are from the reference particle. To calculate
the entropy of the data, {\revision I applied the algorithm in the joint
space of possible parameters and datasets}, i.e., distribution of reference
particles followed Equation~\ref{eqn:joint_prior}.
For the distance function, which helps the NS particles approach the
reference particle, I used a
simple Euclidean metric in the space of datasets:
\begin{align}
d\left(\x; \xref\right) &= \sqrt{
        \sum_{i=1}^{100}\left(x_i - x_{{\rm ref}, i}\right)^2}.
\end{align}
Since the distance function refers only to the data $\boldsymbol{x}$
and not the parameter $\mu$, the sampling is effectively done in the
space of possible datasets only --- the parameter $\mu$ can be thought of
as merely a
latent variable allowing $p(\boldsymbol{x})$ to be explored conveniently.
}

@@

However, more significantly, the above method
would not work if we want $I(\x; \d)$ for
a single parameter or a subset of them, rather than for all of the parameters.
{\revision
The prior-to-posterior KL divergence measured by standard NS is the
amount learned about {\em all} parameters, including nuisance parameters}.
Therefore, I adopted the third formula as the best way to compute $I(\x; \d)$.
In the sinusoidal example, the parameter of interest is the period, whereas
the model will contain other parameters as well.
Since the period will have its prior specified explicitly,
$H(\x)$ will be available analytically. I then use the algorithm to
obtain $H(\x | \d)$ as follows:
\begin{align}
H(\x | \d) &= \int p(\d_{\rm ref}) H(\x | \d=\d_{\rm ref}) \, d\d_{\rm ref}.
\end{align}
This is an expected value over datasets, of the entropy of the posterior given
each dataset. I ran the algorithm by generating reference particles from
$p(\x, \d)$, i.e., generating parameter values from the prior, along with
corresponding simulated datasets. Given each simulated dataset, I used NS
to compute an unbiased estimate of $H(\x | \d=\d_{\rm ref})$, by evolving
a set of particles (initially from the posterior) towards the reference particle.
The role of the ``prior'' in the NS is actually taken by the posterior, and the
``likelihood'' was minus a distance function defined in terms of the
parameter(s) of interest.

@@

Letting $\tau = \log_{10} T$, and treating
$A$ and $\phi$ are nuisance parameters, I calculated the conditional entropy
$H(\tau | \boldsymbol{Y})$ using the algorithm with a tolerance of $r = 10^{-5}$.
{\revision The distance function was just the absolute difference between
$\tau$ and $\tau_{\rm ref}$.}

The result was
\begin{align}
\left<\depth\right> &= 5.379 \pm 0.038
\end{align}
Converting to a differential entropy by adding {\revision $\ln (2r) = -10.820$}, the conditional
entropy is
\begin{align}
H(\tau | \boldsymbol{Y}) &= -5.441 \pm 0.038 \textnormal{ nats}.
\end{align}
Since the Uniform(0,1) prior for $\tau$ has a differential entropy of zero, the
mutual information is $I(\tau; \boldsymbol{Y}) = H(\tau) - H(\tau | \boldsymbol{Y}) = 5.441 \pm 0.038$
nats.

@@

{\revision
The current implementation of the algorithm (see Appendix~\ref{sec:software})
is in C++ and has not been parallelised. Given the high computational
costs, parallelisation should be attempted. This could follow
the method of \citep{HENDERSON201784}, or alternatively, different
threads could work on different reference particles simultaneously.}
